experDesign: helping performing experiments on batches

### 2021-04-21 README.Rmd
Update README and NEWS

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/experDesign)](https://CRAN.R-project.org/package=experDesign)
[![R build status](https://github.com/llrs/experDesign/workflows/R-CMD-check/badge.svg)](https://github.com/llrs/experDesign/actions?workflow=R-CMD-check)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/llrs/experDesign?branch=master&svg=true)](https://ci.appveyor.com/project/llrs/experDesign)
[![Travis build status](https://travis-ci.org/llrs/experDesign.svg?branch=master)](https://travis-ci.org/llrs/experDesign)
[![Coverage status](https://codecov.io/gh/llrs/experDesign/branch/master/graph/badge.svg)](https://codecov.io/github/llrs/experDesign?branch=master)
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://www.tidyverse.org/lifecycle/#stable)
[![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
<!-- badges: end -->

The goal of experDesign is to help you decide which samples go in which batch, 
reducing the potential batch bias when analyzing.

@@

Imagine you have some samples already collected and you want to distributed them in batches:
```{r experDesign_setup}
library("experDesign")
metadata <- expand.grid(height = seq(60, 80, 5), 
                        weight = seq(100, 300, 50),
                        sex = c("Male","Female"))
head(metadata, 15)
```

If you block incorrectly and end up with a group in a single batch we will end up with batch effect.
In order to avoid this `design` helps you assign each sample to a batch (in this case each batch has 24 samples at most). First we can explore the number of samples and the number of batches:

```{r size}
size_data <- nrow(metadata)
size_batch <- 24
(batches <- optimum_batches(size_data, size_batch))
# So now the best number of samples for each batch is less than the available
(size <- optimum_subset(size_data, batches))
# The distribution of samples per batch
sizes_batches(size_data, size, batches)
```

Note that instead of using a whole batch and then leave a single sample on the third distributes all the samples in the three batches that will be needed.
We can directly look for the distribution of the samples given our max number of samples per batch:

```{r design}
d <- design(metadata, size_batch)
# It is a list but we can convert it to a vector with:
batch_names(d)
```

Naively one would either fill some batches fully or distribute them not evenly 
(the  first 17 packages together, the next 17 and so on). This solution ensures 
that the data is randomized. For more random distribution you can increase the number of iterations performed to calculate this distribution.

If you need space for replicates to control for batch effect you can use:

```{r replicates}
r <- replicates(metadata, size_batch, 5)
lengths(r)
r
```

Which seeks as controls the most diverse values and adds them to the samples 
distribution. Note that if the sample is already present on that batch is not added again, that's why the number of samples per batch is different from the design without replicates.

>>>>

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/experDesign)](https://CRAN.R-project.org/package=experDesign)
[![R build status](https://github.com/llrs/experDesign/workflows/R-CMD-check/badge.svg)](https://github.com/llrs/experDesign/actions?workflow=R-CMD-check)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/llrs/experDesign?branch=master&svg=true)](https://ci.appveyor.com/project/llrs/experDesign)
[![Travis build status](https://travis-ci.org/llrs/experDesign.svg?branch=master)](https://travis-ci.org/llrs/experDesign)
[![Coverage status](https://codecov.io/gh/llrs/experDesign/branch/master/graph/badge.svg)](https://codecov.io/github/llrs/experDesign?branch=master)
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
[![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
<!-- badges: end -->

The goal of experDesign is to help you decide which samples go in which batch, 
reducing the potential batch bias before performing an experiment.
It provides three main functions :

* `design()`: Randomize the samples according to their variables.
* `replicates()`: Selects some samples for replicates and randomizes the samples.
* `spatial()`: Randomize the samples on a spatial grid.

@@

We can use the survey dataset for the examples:

```{r show}
library("experDesign")
data(survey, package = "MASS") 
head(survey)
```

The dataset has numeric, categorical values and some `NA`'s value.

# Picking samples for each batch

Imagine that we can only work in groups of 70, and we want to randomize by Sex, 
Smoke, Age, and by writing hand.  
There are `r choose(237, 70)` combinations some of them would be have in a 
single experiment all the right handed students. We could measure all these combinations
but we can try to find an optimum value.

```{r design, fig.show='hold'}
# To reduce the variables used:
omit <- c("Wr.Hnd", "NW.Hnd", "Fold", "Pulse", "Clap", "Exer", "Height", "M.I")
(keep <- colnames(survey)[!colnames(survey) %in% omit])
head(survey[, keep])

# Looking for groups at most of 70 samples.
index <- design(pheno = survey, size_subset = 70, omit = omit)
index
```

We can transform then into a vector to append to the file or to pass to the lab mate with:

```{r batch_names}
head(batch_names(index))
```
